\textbf{ Mapping stellar surfaces\\ I: Degeneracies in the rotational light curve problem }

### 2021-01-26 paper1/ms.tex
misc updates, finished paper 2

For rotational light curves,
the vast majority of the surface modes lie at least partly in the null
space. To show this, we will make use of the fact that we can express
the vector of $K$ observed fluxes $\mathbf{f}$ (i.e., the light curve)
as a linear operation on the vector of $N$
spherical harmonic coefficients $\mathbf{y}$ \citep{Luger2019}:
%
\begin{align}
    \label{eq:fAy}
    \mathbf{f} = \mathbf{A} \, \mathbf{y}
    \quad,
\end{align}
%
where $\mathbf{A}$ is the $(K \times N)$ \emph{design matrix} of the transformation, whose columns
describe how each of the $N$ components in the spherical harmonic basis contribute
to each of the $K$ points in the light curve.%
\footnote{%
    For rotational light curves, the rows of $\mathbf{A}$ are given by the
    quantity $\mathbf{r}^\top \mathbf{A}_1 \mathbf{R}$ in Equation~(18) of
    \citet{Luger2019}, where $\mathbf{r}^\top$ is a vector of disk-integrated
    intensities, $\mathbf{A}_1$ is a change of basis matrix, and $\mathbf{R}$
    is a spherical harmonic rotation matrix that depends on the stellar inclination
    and the rotational phase of the star. Refer to that paper for more details.
}
%
Even though we are explicitly choosing the spherical harmonics as the basis in which
we describe the stellar surface, Equation~(\ref{eq:fAy}) is quite general and applies to
\emph{any} basis that is linearly related to the flux. For instance, $\mathbf{y}$
could instead describe the intensities in each of the $N$ pixels of a gridded stellar
surface, in which case $\mathbf{A}$ would be the matrix of pixel visibilities that describe
how to sum each of the pixels to obtain the observed light curve.

The size of the null space is called the \emph{nullity}, and it is equal to
$N - R$, where $N$ is once again the number of coefficients describing the
stellar surface and $R$ is the \emph{rank} of the flux operator $\mathbf{A}$.
The rank $R$ is the number of linearly independent columns in $\mathbf{A}$, which
can be computed numerically using any standard linear algebra package. It is equal
to the number of independent signals (or components) that can be measured given
an observation of $\mathbf{f}$.

@@

Figure~\ref{fig:rank} shows the rank and nullity of the flux operator
as a function of the resolution of the surface map (quantified as the
spherical harmonic degree $l$ of the expansion). The orange
curve is the number of spherical harmonic coefficients needed to
represent a surface map up to degree $l$, and is equal to $N = (l + 1)^2$.
The blue line shows the rank of the flux operator $\mathbf{A}$ as a function
of $l$, which scales as $R \approx 2l + 1$. The nullity is
simply the difference between $N$ and $R$.

@@

\subsection{Decomposition of the flux operator}
%
\label{sec:svd}
%
In Appendix~\ref{sec:app-svd} we show that it is straightforward to
construct linear operators $\mathbf{P}$ and $\mathbf{N}$ from the
design matrix $\mathbf{A}$ such that
%
\begin{align}
    \mathbf{y}_\bullet & = \mathbf{P} \, \mathbf{y}
\end{align}
%
is the component of the surface map in the \emph{preimage} and
%
\begin{align}
    \mathbf{y}_\circ & = \mathbf{N} \, \mathbf{y}
\end{align}
%
is the component of the map in the \emph{null space}.
%
The \emph{preimage operator} $\mathbf{P}$
transforms a vector $\mathbf{y}$ in the surface map basis in such a way
that it preserves the information in $\mathbf{y}$ that gets mapped
onto the light curve $\mathbf{f}$ via $\mathbf{A}$ (the \emph{preimage}) and discards the rest. The
\emph{null space operator} $\mathbf{N}$ does the opposite: it preserves only
the information in $\mathbf{y}$ that gets mapped onto the zero
vector via $\mathbf{A}$ (the \emph{null space}).
%
In other words,
the $\mathbf{P}$ and $\mathbf{N}$ operators reveal the
components of the surface map that contribute to the
light curve ($\mathbf{y}_\bullet$) and the components that don't ($\mathbf{y}_\circ$).
The vector $\mathbf{y}_\bullet$ represents all the information that can be
learned from a stellar light curve, while $\mathbf{y}_\circ$ represents all the
information that cannot.

@@

It is instructive to visualize these components in
an actual surface mapping exercise. Figure~\ref{fig:nullspace_preimage}
shows the decomposition of four hypothetical surfaces (left
column) into preimage (center) and null space (right)
components under the flux operator $\mathbf{A}$, which we compute for
definiteness at an inclination of $I = 60^\circ$ over a full rotation.
The light curves corresponding to each of these surfaces are also shown.
Note that both the true map and its associated light curve are simply equal to
the sum of the preimage and null space components (Appendix~\ref{sec:app-svd}).
%
As expected, all of the information in the light curve comes
from the preimage, and the null space contributes exactly zero
flux at all phases. However, most of the information about the
\emph{surface} is stuck in the null space!

@@

Let us therefore consider Figure~\ref{fig:nullspace_preimage} in the
context of an inference exercise. One can equivalently think of
the preimage as the solution to a least squares inference problem
in which a completely uninformative prior%
\footnote{%
    More specifically, a zero-mean, infinite variance Gaussian prior on the coefficients of $\mathbf{y}$.
}
is placed on the spherical harmonic
coefficients. In this case, the least squares solution is given by
%
\begin{proof}{test_svd_lstsq}
    \label{eq:inference}
    \hat{\mathbf{y}}
    &=
    \lim_{\lambda \rightarrow 0} \left( \mathbf{A}^\top \mathbf{A} + \lambda \mathbf{I}\right)^{-1} \mathbf{A}^\top \mathbf{f}
    \nonumber\\
    &= \mathbf{y}_\bullet
    \quad.
\end{proof}
%
In this sense, $\mathbf{y}_\bullet$ represents our knowledge about the
surface of a star after an observation if we have no prior information
whatsoever on $\mathbf{y}$.
\emph{Any other information about the surface must come from the prior.}%

@@

Another recent paper relevant to our work is that of \citet{Basri2020}, who
investigated the information content of stellar light curves, exploring what
can and cannot be learned about star spot configurations from individual
light curves. That paper strongly urges against the common practice
of interpreting
light curves with one or two dips as originating from one or two spots,
respectively, a point we strongly agree with (c.f. our Figure~\ref{fig:degeneracies}).
It also reinforces our point about the additional degeneracies introduced
by the unknown baseline inherent to single-band photometry.
%

@@

    \mathbf{A} = \mathbf{U} \, \mathbf{S} \, \mathbf{V}^\top

@@

\emph{singular vectors} of $\mathbf{A}$, and the entries along the
diagonal of $\mathbf{S}$ are the corresponding \emph{singular values}. If $\mathbf{A}$

@@

We may then express the decomposition of $\mathbf{A}$ as

@@

    \mathbf{A} =

@@

    \mathbf{f} & = \mathbf{A} \, \mathbf{y}

@@

        \mathbf{f} = \mathbf{A} \, \mathbf{y}_\bullet

@@

where the fact that $\mathbf{U}_\bullet \, \mathbf{S}_\bullet \, \mathbf{V}_\bullet^\top = \mathbf{A}$

>>>>

For rotational light curves,
the vast majority of the surface modes lie at least partly in the null
space. To show this, we will make use of the fact that we can express
the vector of $K$ observed fluxes $\mathbf{f}$ (i.e., the light curve)
as a linear operation on the vector of $N$
spherical harmonic coefficients $\mathbf{y}$ \citep{Luger2019}:
%
\begin{align}
    \label{eq:fAy}
    \mathbf{f} = \mathbf{1} + \pmb{\mathcal{A}} \, \mathbf{y}
    \quad,
\end{align}
%
where $\pmb{\mathcal{A}}$ is the $(K \times N)$ \emph{design matrix} of the transformation, whose columns
describe how each of the $N$ components in the spherical harmonic basis contribute
to each of the $K$ points in the light curve.%
\footnote{%
    For rotational light curves, the rows of $\pmb{\mathcal{A}}$ are given by the
    quantity $\mathbf{r}^\top \pmb{\mathcal{A}}_1 \mathbf{R}$ in Equation~(18) of
    \citet{Luger2019}, where $\mathbf{r}^\top$ is a vector of disk-integrated
    intensities, $\pmb{\mathcal{A}}_1$ is a change of basis matrix, and $\mathbf{R}$
    is a spherical harmonic rotation matrix that depends on the stellar inclination
    and the rotational phase of the star. In Equation~(\ref{eq:fAy}) we explicitly
    add a vector of ones to enforce a unit baseline for spotless stars.
    Refer to \citet{Luger2019} for more details.
}
%
Even though we are explicitly choosing the spherical harmonics as the basis in which
we describe the stellar surface, Equation~(\ref{eq:fAy}) is quite general and applies to
\emph{any} basis that is linearly related to the flux. For instance, $\mathbf{y}$
could instead describe the intensity deficits in each of the $N$ pixels of a gridded stellar
surface, in which case $\pmb{\mathcal{A}}$ would be the matrix of pixel visibilities that describe
how to sum each of the pixels to obtain the observed light curve.

The size of the null space is called the \emph{nullity}, and it is equal to
$N - R$, where $N$ is once again the number of coefficients describing the
stellar surface and $R$ is the \emph{rank} of the flux operator $\pmb{\mathcal{A}}$.
The rank $R$ is the number of linearly independent columns in $\pmb{\mathcal{A}}$, which
can be computed numerically using any standard linear algebra package. It is equal
to the number of independent signals (or components) that can be measured given
an observation of $\mathbf{f}$.

@@

Figure~\ref{fig:rank} shows the rank and nullity of the flux operator
as a function of the resolution of the surface map (quantified as the
spherical harmonic degree $l$ of the expansion). The orange
curve is the number of spherical harmonic coefficients needed to
represent a surface map up to degree $l$, and is equal to $N = (l + 1)^2$.
The blue line shows the rank of the flux operator $\pmb{\mathcal{A}}$ as a function
of $l$, which scales as $R \approx 2l + 1$. The nullity is
simply the difference between $N$ and $R$.

@@

\subsection{Decomposition of the flux operator}
%
\label{sec:svd}
%
In Appendix~\ref{sec:app-svd} we show that it is straightforward to
construct linear operators $\mathbf{P}$ and $\mathbf{N}$ from the
design matrix $\pmb{\mathcal{A}}$ such that
%
\begin{align}
    \mathbf{y}_\bullet & = \mathbf{P} \, \mathbf{y}
\end{align}
%
is the component of the surface map in the \emph{preimage} and
%
\begin{align}
    \mathbf{y}_\circ & = \mathbf{N} \, \mathbf{y}
\end{align}
%
is the component of the map in the \emph{null space}.
%
The \emph{preimage operator} $\mathbf{P}$
transforms a vector $\mathbf{y}$ in the surface map basis in such a way
that it preserves the information in $\mathbf{y}$ that gets mapped
onto the light curve $\mathbf{f}$ via $\pmb{\mathcal{A}}$ (the \emph{preimage}) and discards the rest. The
\emph{null space operator} $\mathbf{N}$ does the opposite: it preserves only
the information in $\mathbf{y}$ that gets mapped onto the zero
vector via $\pmb{\mathcal{A}}$ (the \emph{null space}).
%
In other words,
the $\mathbf{P}$ and $\mathbf{N}$ operators reveal the
components of the surface map that contribute to the
light curve ($\mathbf{y}_\bullet$) and the components that don't ($\mathbf{y}_\circ$).
The vector $\mathbf{y}_\bullet$ represents all the information that can be
learned from a stellar light curve, while $\mathbf{y}_\circ$ represents all the
information that cannot.

@@

It is instructive to visualize these components in
an actual surface mapping exercise. Figure~\ref{fig:nullspace_preimage}
shows the decomposition of four hypothetical surfaces (left
column) into preimage (center) and null space (right)
components under the flux operator $\pmb{\mathcal{A}}$, which we compute for
definiteness at an inclination of $I = 60^\circ$ over a full rotation.
The light curves corresponding to each of these surfaces are also shown.
Note that both the true map and its associated light curve are simply equal to
the sum of the preimage and null space components (Appendix~\ref{sec:app-svd}).
%
As expected, all of the information in the light curve comes
from the preimage, and the null space contributes exactly zero
flux at all phases. However, most of the information about the
\emph{surface} is stuck in the null space!

@@

Let us therefore consider Figure~\ref{fig:nullspace_preimage} in the
context of an inference exercise. One can equivalently think of
the preimage as the solution to a least squares inference problem
in which a completely uninformative prior%
\footnote{%
    More specifically, a zero-mean, infinite variance Gaussian prior on the coefficients of $\mathbf{y}$.
}
is placed on the spherical harmonic
coefficients. In this case, the least squares solution is given by
%
\begin{proof}{test_svd_lstsq}
    \label{eq:inference}
    \hat{\mathbf{y}}
    &=
    \lim_{\lambda \rightarrow 0} \left( \pmb{\mathcal{A}}^\top \pmb{\mathcal{A}} + \lambda \mathbf{I}\right)^{-1} \pmb{\mathcal{A}}^\top (\mathbf{f} - \mathbf{1})
    \nonumber\\
    &= \mathbf{y}_\bullet
    \quad.
\end{proof}
%
In this sense, $\mathbf{y}_\bullet$ represents our knowledge about the
surface of a star after an observation if we have no prior information
whatsoever on $\mathbf{y}$.
\emph{Any other information about the surface must come from the prior.}%

@@

Another recent paper relevant to our work is that of \citet{Basri2020}, who
investigated the information content of stellar light curves, exploring what
can and cannot be learned about star spot configurations from individual
light curves. That paper strongly urges against the common practice
of interpreting
light curves with one or two dips as originating from one or two spots,
respectively, a point we strongly agree with (see our Figure~\ref{fig:degeneracies}).
It also reinforces our point about the additional degeneracies introduced
by the unknown baseline inherent to single-band photometry.
%

@@

    \pmb{\mathcal{A}} = \mathbf{U} \, \mathbf{S} \, \mathbf{V}^\top

@@

\emph{singular vectors} of $\pmb{\mathcal{A}}$, and the entries along the
diagonal of $\mathbf{S}$ are the corresponding \emph{singular values}. If $\pmb{\mathcal{A}}$

@@

We may then express the decomposition of $\pmb{\mathcal{A}}$ as

@@

    \pmb{\mathcal{A}} =

@@

    \mathbf{f} - \mathbf{1} & = \pmb{\mathcal{A}} \, \mathbf{y}

@@

        \mathbf{f} = \mathbf{1} + \pmb{\mathcal{A}} \, \mathbf{y}_\bullet

@@

where the fact that $\mathbf{U}_\bullet \, \mathbf{S}_\bullet \, \mathbf{V}_\bullet^\top = \pmb{\mathcal{A}}$
